Map{R}educe in {MPI} for Large-scale Graph Algorithms

### 2010-05-03 paper/outcore.tex
More edits; consistent notation.
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@602 3fc257bb-8550-0410-81b3-a5c103da83f7

Since the MapReduce paradigm was designed to enable processing of
extremely large data sets, the MR-MPI library also allows for
``out-of-core'' processing, which is triggered when the KV or KMV
pairs owned by oen or more processors do not fit in its local memory.
When this occurs, a processor writes one or more temporary files to
disk, containing KV or KMV pairs, and reads them back in when
required.  Depending on the parallel machine's hardware configuration,
these files can reside on disks local to each processor, on the front
end (typically a NSF-mounted file system for a parallel machine), or
on a parallel file system/disk array.

When a user program creates a MapReduce object, a ``pagesize'' can be
specified, which defaults to 64 Mbytes.  As described below, each
MR-MPI operation, is constrained to use no more than a handful of
these pages.  The {\it pagesize} setting can be as small as 1 Mbyte or
as large as desired, though the user should insure the allocated pages
fit in physical memory, else the MR-MPI library may allocate slow
virtual memory.  The {\it pagesize} is also the typical size of
individual reads and writes to the temporary disk files; hence a
reasonable {\it pagesize} insures good I/O performance.

@@

Performing an our-of-core {\it collate()} operation is more complex.
Recall that the operation occurs in two stages.  First, keys are
hashed to ``owning'' processors and the KV datums are communicated to
new processors in an all-to-all fashion.  In out-of-core mode, this is
done one page of KV data at a time.  Each processor reads in a page of
KV pairs, the communication pattern for the datums in that page is
determined (who needs what datums), and all-to-all communication is
performed.  Each processor allocates a two-page chunk of memory to
receive incoming KV pairs.  On average each processor should receive
one page of KV pairs; the two-page allocation allows for some
load-imbalance.  If the KV pairs are distributed unevenly so that this
limit is exceeded, then the all-to-all communication is performed in
smaller, multiple passes until the full page contributed by every
processor has been communicated.  Performing an MPI\_Alltoall(), or
using the custom all-to-all routines provided by the MR-MPI library,
requires allocation of auxiliary arrays that store processor indices,
datum pointers, and datum lengths.  For the case of many tiny KV
datums, the total number of memory pages required to perform the
all-to-all communication, including the source and target KV pages is
at most 7.

When the communication stage of the {\it collate()} operation is
complete, each processor now has a set of KV pages, stored in an
out-of-core KV file, that need to be reorganized into a set of KMV
pages, also stored in new out-of-core KMV file.  In principle, to
create one page of KMV pairs, would require all KV pages be scanned,
to find all the keys which contribute values to that page.  Doing this
for each output page of KMV pairs could be prohibitively expensive.  A
related issue is that, as described in the previous section, a hash
table is needed to match new keys with previously encountered keys.
But there is no guarantee that the hash table itself will not grow
arbitrarily large for data sets with many unique keys.  We need an
algorithm for generating KMV pairs which operates within a small
number of memory pages and performs a minimal number of passes through
the KV disk file.  An algorithm that meets these goals is diagrammed
in Figure \ref{fig:collate}; it reads the KV pages at most 4 times, and
writes out new KV or KMV pages at most 3 times.  It also uses a
finite-size in-memory hash table, which stores unique keys as they are
encountered while looping over the KV pages, as well as auxiliary
information needed to construct the output pages of KMV pairs.

@@

\item (Pass 3) The unique keys in the HT for one partition are
scanned.  The associated values may create KMV pairs which will span
$M$ memory pages.  If $M > 1$ the associated partition file of KV
pairs is read again, and each KV pair is assigned to one of $M$
smaller ``set'' files, represented by KVs in the figure.  Each set
file contains the KV pairs which will contribute to the KMV pairs that
will populate one page of KMV output.  If a single KMV pair spans
multiple pages, because it contains a very large number of values,
then the corresponding KV pairs will still be written to a single set
file.  Eventually, this pass reads all partition files, entailing both
a read and write of all KV pairs.

@@

In summary, this data reorganization for the {\it collate()} operation
is somewhat complex, but requires only a small, constant number of
passes through the data.  The KV datums are read from disk at most 4
times, and written to disk 3 times.  The first two write passes
reorganize KV pairs into partition and set files; the final write pass
creates the KMV data set.  Depending on the size and characteristics
of the KV datums (e.g. how many unique keys it contains), some of
these passes may not be needed.  By contrast, a full out-of-core sort
of the KV pairs, performed via a merge sort as discussed below,
requires $O(\log{M})$ read/write passes through the KV data, where $M$
is the number of pages of KV pairs, which can be a large number for
big data sets.

The memory page requirements for the above algorithm are as follows.
Two contiguous pages of memory are used for the hash table.
Intermediate passes 2 and 3 can require numerous partition or set
files to be opened simultaneously and written to.  To avoid small
writes of individual KV datums, a buffer of at least 16K bytes is
allocated for each file.  The precise number of simultaneously open
files is difficult to bound, but typically one or two additional pages
memory suffice for all needed buffers.  Thus the total pages required
for the data reorganization stage of the {\it collate()} operation is
less than the 7 used by the all-to-all communication stage.

Other MR-MPI library calls discussed in the previous section, such as
{\it compress()}, {\it convert()}, {\it collapse()}, {\it gather()},
etc, can also operate in out-of-core mode, typically with one pass
through the KV or KMV data and the use of one or two memory pages.
Sorting KV pairs, by key or value, is an exception.  An out-of-core
merge sort is performed as follows.  Two pages of KV datums are read
from disk.  Each is sorted using the C-library quicksort() function,
as discussed in the previous section.  The two pages are then scanned
and merged into a new file which is written to disk as its associated
memory page fills up.  This process requires 5 memory pages, which
includes vectors of KV datum pointers and lengths needed by the
in-memory quicksort() operation.  Once all pairs of pages have been
merged, the sort continues in a recursive fashion, merging pairs of
files into a new third file, without the need to perform additional
in-memory sorts.  Thus the overall memory requirement is 5 pages.  The
number of read/write passes through the KV data set for the merge sort
is $O(\log{M})$, where $M$ is the number of pages of KV pairs.

>>>>

Since the MapReduce paradigm was designed to enable processing of
extremely large data sets, the MR-MPI library also allows for
``out-of-core'' processing, which is triggered when the KV or KMV
pairs owned by one or more processors do not fit in local memory.
When this occurs, a processor writes one or more temporary files to
disk, containing KV or KMV pairs, and reads them back in when
required.  Depending on the parallel machine's hardware configuration,
these files can reside on disks local to each processor, on the front
end (typically a NSF-mounted file system for a parallel machine), or
on a parallel file system/disk array.

When a user program creates a MapReduce object, a ``pagesize'' can be
specified, which defaults to 64 Mbytes.  As described below, each
MR-MPI operation is constrained to use no more than a handful of
these pages.  The {\it pagesize} setting can be as small as 1 Mbyte or
as large as desired, though the user should ensure the allocated pages
fit in physical memory; otherwise, the MR-MPI library may allocate slow
virtual memory.  The {\it pagesize} is also the typical size of
individual reads and writes to the temporary disk files; hence a
reasonable {\it pagesize} ensures good I/O performance.

@@

Performing an our-of-core {\it collate()} operation is more complex.
Recall that the operation occurs in two stages.  First, keys are
hashed to ``owning'' processors and the KV datums are communicated to
new processors in an all-to-all fashion.  In out-of-core mode, this 
operation is
performed on one page of KV data at a time.  Each processor reads in a page of
KV pairs, the communication pattern (which processors will receive 
which datums in that page) is
determined, and all-to-all communication is
performed.  Each processor allocates a two-page chunk of memory to
receive incoming KV pairs.  On average each processor should receive
one page of KV pairs; the two-page allocation allows for some
load-imbalance.  If the KV pairs are distributed unevenly so that this
limit is exceeded, then the all-to-all communication is performed in
smaller, multiple passes until the full page contributed by every
processor has been communicated.  Performing an MPI\_Alltoall(), or
using the custom all-to-all routines provided by the MR-MPI library,
requires allocation of auxiliary arrays that store processor indices,
datum pointers, and datum lengths.  For the case of many tiny KV
datums, the total number of memory pages required to perform the
all-to-all communication, including the source and target KV pages is
at most seven.

When the communication stage of the {\it collate()} operation is
complete, each processor has a set of KV pages, stored in an
out-of-core KV file, that need to be reorganized into a set of KMV
pages, stored in a new out-of-core KMV file.  In principle, 
creating one page of KMV pairs would require all KV pages be scanned
to find all the keys that contribute values to that page.  Doing this
for each output page of KMV pairs could be prohibitively expensive.  A
related issue is that, as described in the previous section, a hash
table is needed to match new keys with previously encountered keys.
But there is no guarantee that the hash table itself will not grow
arbitrarily large for data sets with many unique keys.  We need an
algorithm for generating KMV pairs that operates within a small
number of memory pages and performs a minimal number of passes through
the KV disk file.  An algorithm that meets these goals is diagrammed
in Figure \ref{fig:collate}; it reads the KV pages at most four times, and
writes out new KV or KMV pages at most three times.  It also uses a
finite-size in-memory hash table, which stores unique keys as they are
encountered while looping over the KV pages, as well as auxiliary
information needed to construct the output pages of KMV pairs.

@@

\item (Pass 3) The unique keys in the HT for one partition are
scanned to determine the total size of KMV for keys in the HT.  
The associated values may create KMV pairs that will span
$M$ memory pages.  If $M > 1$ the associated partition file of KV
pairs is read again, and each KV pair is assigned to one of $M$
smaller ``set'' files, represented by KVs in the figure.  Each set
file contains the KV pairs that will contribute to the KMV pairs that
will populate one page of KMV output.  If a single KMV pair spans
multiple pages, because it contains a very large number of values,
then the corresponding KV pairs will still be written to a single set
file.  Eventually, this pass reads all partition files, entailing both
a read and write of all KV pairs.

@@

In summary, this data reorganization for the {\it collate()} operation
is somewhat complex, but requires only a small, constant number of
passes through the data.  The KV datums are read from disk at most four
times, and written to disk three times.  The first two write passes
reorganize KV pairs into partition and set files; the final write pass
creates the KMV data set.  Depending on the size and characteristics
of the KV datums (e.g., the number of unique keys), some of
these passes may not be needed.  By contrast, a full out-of-core sort
of the KV pairs, performed via a merge sort as discussed below,
requires $O(\log{M})$ read/write passes through the KV data, where $M$
is the number of pages of KV pairs, which can be a large number for
big data sets.

The memory page requirements for the above algorithm are as follows.
Two contiguous pages of memory are used for the hash table.
Intermediate passes 2 and 3 can require numerous partition or set
files to be opened simultaneously and written to.  To avoid small
writes of individual KV datums, a buffer of at least 16K bytes is
allocated for each file.  The precise number of simultaneously open
files is difficult to bound, but typically one or two additional pages'
memory suffice for all needed buffers.  Thus the total number of pages required
for the data reorganization stage of the {\it collate()} operation is
less than the seven used by the all-to-all communication stage.

Other MR-MPI library calls discussed in the previous section, such as
{\it compress()}, {\it convert()}, {\it collapse()}, and {\it gather()},
can also operate in out-of-core mode, typically with one pass
through the KV or KMV data and the use of one or two memory pages.
Sorting KV pairs, by key or value, is an exception.  An out-of-core
merge sort is performed as follows.  Two pages of KV datums are read
from disk.  Each is sorted using the C-library quicksort() function,
as discussed in the previous section.  The two pages are then scanned
and merged into a new file which is written to disk as its associated
memory page fills up.  This process requires five memory pages, which
includes vectors of KV datum pointers and lengths needed by the
in-memory quicksort() operation.  Once all pairs of pages have been
merged, the sort continues in a recursive fashion, merging pairs of
files into a new third file, without the need to perform additional
in-memory sorts.  Thus the overall memory requirement is five pages.  The
number of read/write passes through the KV data set for the merge sort
is $O(\log{M})$, where $M$ is the number of pages of KV pairs.
